The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-07-18 paper.qmd
wrote the ignoring uncertainty section

# load Libraries
library(tidyverse)
library(biscale)
library(RColorBrewer)
library(scales)
library(sf)
library(ggrepel)
library(urbnmapr)

```
# Background

@@

# Ignoring uncertainty

## Choropleth Map

## Why include uncertainty?

# Visualising Uncertainty as a Variable
## Bivariate Maps
## Signal Supression
## Uncertainty without signal supression

# Evaluating uncertainty
# Current methods
# Testing signal supression

# ------------------- BELOW HERE IS UNSORTED ------------------------
# Why is uncertainty special?
It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why*. A visualisation is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic. Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, median, maximum, or of any other statistic, however this field *does* implicitly treat them different. There is no such thing as "maximum visualiation" or "median visualisation" so there *must* be something about uncertainty that causes authors to believe it is special. Here we seek to answer this question.

### The purpose of uncertainty visualisations
Uncertainty visualisation papers usually provide one of two primary justifications for their existence. The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty. Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019]. The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate. Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].

Uncertainty, acting as it is intended, is transparency. There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors. These secondary benefits are, however, *not* the immediate goal of uncertainty. The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.

## What is signal-supression?
The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.

Interviews with experts in visualisation and statistics often back up "signal suppression" as the primary motivation for uncertainty visualisation. Failling to communicate uncertainty is often considered to be akin to fraud or lying [@Hullman2020a; @Manski2020] indicating that uncertainty should exist to prevent undue certainty in visual inference. The relationship between signal supression and uncertainty visualisation is made more explicity by other authors [@Griethe2006]. Therefore uncertainty acts as a form of statistical "hedging" for signals found in an analysis. This aligns nicely with our understanding of uncertainty as a by product of inference. The variable we are trying to perform inference on is the signal, while the while the visualisation of noise that prevents the signal from being read is the uncertainty. It is nice that the connection to inference easily extends from uncertainty to uncertainty visualisation, however many of the confusions extend as well.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-maps
#| fig-cap: "A visualisation of a typical choropleth map, as well as three other maps that are used to display uncertainty on the choropleth map. There is a high variance and low variance example for each type of map to show how well the technique supresses signal. Each map was created using the same data with the same base palette. At first glance, the high uncertainty bivariate and VSUP maps just look like maps with a low saturation colour palette, rather than map with high uncertainty. This visualisation makes it clear that supression methods that plot uncertainty to a second axis, such as hue, make uncertainty appear as a second variable, rather than a signal supression on our estimate. "
#| fig-subcap: 
#|   - "Choropleth Map: Low Variance"
#|   - "Choropleth Map: High Variance"
#|   - "Choropleth Palette"
#|   - "Bivariate Map: Low Variance"
#|   - "Bivariate Map: High Variance"
#|   - "Bivariate Palette"
#|   - "VSUP Map: Low Variance"
#|   - "VSUP Map: High Variance"
#|   - "VSU Palette"
#|   - "Pixel Map: Low Variance"
#|   - "Pixel Map: High Variance"
#|   - "Pixel Palette"
#| layout-ncol: 3

@@

p1b <- p1a %+% filter(my_map_data, variance_class=="highvar")
# Bivariate Map
# Make bivariate palette
# Function to devalue by a certain amount
colsupress <- function(basecols, hue=1, sat=1, val=1) {
    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
}

@@

# Display plots and palettes
p1a
p1b
show_col(basecols, ncol=8, labels=FALSE)
p2a
p2b
show_col(bivariatepal, ncol=8, labels=FALSE)
p3a
p3b
show_col(VSUP, ncol=8, labels=FALSE)
p4a
p4b
show_col(basecols, ncol=8, labels=FALSE)
```

>>>>

# load Libraries
library(tidyverse)
library(biscale)
library(RColorBrewer)
library(scales)
library(sf)
library(ggrepel)
library(urbnmapr)
```
# Background

@@

# Ignoring uncertainty
Before we look at different cases where uncertainty is included, it is beneficial for us to look at cases where it is not. After all, why include uncertainty at all?

@fig-choropleth depicts a choropleth map of the counties of Iowa. Each of these counties are colored according to an estimate of average daily temperature that was simulated so that the values followed a clear spatial trend (hot in the middle of the map, and cold on the outside). The variance of these estimates were simulated such that a hypothesis test would indicate the existence of a spatial trend in the low variance map, while the trend in the high variance map should be indistinguishable from noise. Is this distinction in validity of the spatial trend clear in in the map? Is the validity of the trend communicated through the visualisation?

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-choropleth
#| fig-cap: "Two choropleth maps that depicts the counties of Iowa where each country coloured acording to a simulated average temperature. Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map. This trend is only technically there in the low variance condition, i.e. if we performed a hypothesis test we would conclude the trend exists. The high variance condition displays a spatial trend that could simply be the result of noise, which means the plot is displaying a false conclusion."

@@

p1b <- p1a %+% filter(my_map_data, variance_class=="highvar")

p1a
p1b
show_col(basecols, ncol=8, labels=FALSE)
```

The two choropleth maps appearing to be identical in @fig-choropleth highlights the need for uncertainty visualisations. Uncertainty visualisation is is required for transparency and this sentiment has been repeated.  Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the "validity" of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018]. Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].

This consensus leads us to understand that uncertainty visualisation is motivated by the need for a sort of "visual hypothesis test". A successful uncertainty visualisation would act as a "statistical hedge" for any inference me make using the graphic. This implies three key criteria for a good uncertainty visualisation:

  1) Reinforce justified conclusions to encourage confidence in results
  2) Prevent unjustified conclusions that are indistinguishable from noise
  3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.

As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. Therefore, we call this approach "signal-supression" as it primarily differentiates itself from from a typical visualisation approach through criteria (2), that is, an uncertainty visualisation should prevent us from drawing unjustified conclusions. We will investigate several approaches for uncertainty visualisation in the following sections and in doing so, discuss how successful each approach is in fulfilling the signal-suppression criteria.

# Visualising Uncertainty as a Variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. You may think "well, I will just add a dimension to my plot that includes uncertainty". This is not a silly choice, but lets look at what happens when we do this.

## Bivariate Map
## Signal Supression

The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.

It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why*. A visualisation is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic. Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, median, maximum, or of any other statistic, however this field *does* implicitly treat them different. There is no such thing as "maximum visualiation" or "median visualisation" so there *must* be something about uncertainty that causes authors to believe it is special. Here we seek to answer this question.

Uncertainty visualisation papers usually provide one of two primary justifications for their existence. The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty. Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019]. The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate. Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].

## Can we have uncertainty without signal supression?
Some authors do not describe uncertainty as a means of transparency, but rather as a variable of importance in of itself [@Blenkinsop2000].

It is not entirely true that uncertainty is only important for transparency, however it is not clear that the "uncertainty" provided in other cases is uncertainty at all. Some authors claim the purpose of uncertainty is not transparency, but rather important information for effective decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. While this can be true, the uncertainty provided for this purpose is *not* the same as the uncertainty provided for transparency. It arguably may not be uncertainty at all. This is easiest to understand with an example.

Let's say my weather prediction app only provides me with the predicted daily rainfall when I am trying to decide if I should bring an umbrella. While my decision will be improved with the inclusion of uncertainty information, that is not because of a special quality of uncertainty, but rather because the uncertainty information can be used to calculate the true statistic of interest, the probability that it will rain at all. If I use uncertainty to improve decision making, it is not *true* uncertainty, rather it is supplementary material that is required to correct for an insufficient statistic. I was provided with was not the statistic required to make the decison.

# Evaluating uncertainty
# Current methods
# Testing signal supression
## Current methods
There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors. These secondary benefits are, however, *not* the immediate goal of uncertainty. The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.

# ------------------- BELOW HERE IS UNSORTED ------------------------
# Why is uncertainty special?

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-maps
#| fig-cap: "A visualisation of a typical choropleth map, as well as three other maps that are used to display uncertainty on the choropleth map. There is a high variance and low variance example for each type of map to show how well the technique supresses signal. Each map was created using the same data with the same base palette. At first glance, the high uncertainty bivariate and VSUP maps just look like maps with a low saturation colour palette, rather than map with high uncertainty. This visualisation makes it clear that supression methods that plot uncertainty to a second axis, such as hue, make uncertainty appear as a second variable, rather than a signal supression on our estimate. "

# Bivariate Map
# Make bivariate palette
# Function to devalue by a certain amount
colsupress <- function(basecols, hue=1, sat=1, val=1) {
    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
}

@@

# Display plots and palettes
p2a
p2b
show_col(bivariatepal, ncol=8, labels=FALSE)
p3a
p3b
show_col(VSUP, ncol=8, labels=FALSE)
p4a
p4b
show_col(basecols, ncol=8, labels=FALSE)
```
